### 2020-11-11 latex/vlp.tex
Incorporating Manfredi's changes to VLP proof

Using this configuration, we will compute the Hamiltonian dynamics of the system.
The cartesian position of the mass at the tip of the pendulum
is given by \(x = (l\sin(q),-l\cos(q))\), while its velocity is
\(\dot{x} = (\dot{l}\sin(q) + l\cos(q)\dot{q}, -\dot{l}\cos(q) + l\sin(q)\dot{q})\).
Computing the kinetic energy \(T\) yields
\[
   T(\mathbf{q},\dot{\mathbf{q}}) = 
   \frac{1}{2}m\norm{\dot{x}}^2 = \frac{1}{2}m\left(\dot{l}^2 + l^2\dot{q}^2\right)
   .
\]
The potential energy \(P\) with respect to the pivot (under a gravitational
acceleration \(g\)) is
\[
   P(\mathbf{q}) = -mgl\cos(q)
   .
\]
Collecting the kinetic energy into a quadratic form, we get the Lagrangian
\[
   \mathcal{L}(\mathbf{q},\dot{\mathbf{q}}) 
   = \frac{1}{2} \dot{\mathbf{q}}\tpose D(\mathbf{q})\dot{\mathbf{q}} - P(\mathbf{q})
   = \frac{1}{2}
   \begin{bmatrix} \dot{q} & \dot{l} \end{bmatrix}
   \begin{bmatrix}
      ml^2 & 0 \\
      0 & m \\
   \end{bmatrix}
   \begin{bmatrix} 
      \dot{q} \\ \dot{l}
   \end{bmatrix}
   + mgl\cos(q)
   .
\]
Computing the conjugate of momenta to \(\mathbf{q}\), we get 
\[
   \mathbf{p} := \begin{bmatrix} p \\ p_l \end{bmatrix} 
   = \begin{bmatrix} ml^2\dot{q} \\ m\dot{l} \end{bmatrix} 
   .
\]
Performing the Legendre transform on \(\mathcal{L}\) and setting
\(M(\mathbf{q}) := D(\mathbf{q})\), \(V(\mathbf{q}) := P(\mathbf{q})\),
we get the Hamiltonian (\ref{eqn:simply-actuated-hamiltonian}) whose
dynamics (\ref{eqn:simply-actuated-dynamics}) resolve to  
\begin{align}\label{eqn:vlp-hamiltonian-with-pl}
   \mathcal{H} &= \frac{1}{2} \begin{bmatrix} p & p_l \end{bmatrix}
      \begin{bmatrix}
         \frac{1}{ml^2}  & 0 \\
         0 & \frac{1}{m}
      \end{bmatrix} \begin{bmatrix} p \\ p_l \end{bmatrix} - mgl\cos(q)
      , \\
     &\begin{cases}
        \dot{q} = \frac{p}{ml^2} \\
        \dot{l} = \frac{p_l}{m} \\
        \dot{p} = -mgl\sin(q) \\
        \dot{p}_l = \frac{p^2}{ml^3} + mg\cos(q) + \tau
        . \\
   \end{cases} \nonumber
\end{align}
The control input is a force \(\tau \in \mathbb{R}\) affecting the dynamics of
\(p_l\), acting colinearly with the rod.
We assume the force does not affect the dynamics of \(p\) in any way -
that is, the control input cannot enact any lateral force on the pendulum.
This makes the VLP into an underactuated mechanical system with degree of
underactuation one. 
It is also a useful assumption because it means \((\mathbf{q},\mathbf{p})\) 
are simply actuated coordinates, which allows us to apply the theory of VNHCs we
developed in Chapter \ref{ch:vnhcs}.

@@

\begin{defn}\label{defn:energy-injection}
   A regular VNHC \(h(\mathbf{q},\mathbf{p}) = 0\) 
   \textit{injects (dissipates) energy} if the constrained dynamics gain (lose)
   energy everywhere on the constraint manifold, except possibly on a set of
   measure zero.
\end{defn}

@@

\begin{thm}\label{thm:vlp-energy-stabilization}
   Define \(\theta := \arctan_2(p,q)\).
   Let \(L : \mathbb{S}^1 \rightarrow [\underline{l},\overline{l}]\) be a
   \(C^2\) function of \(\theta\).
   A regular VNHC of the form 
   \(h(\mathbf{q},\mathbf{p}) = l - L(\theta)\) injects energy into the VLP
   if there exists 
   \(l_\text{avg} \in [\underline{l},\overline{l}]\) such that 
   \begin{equation}\label{eqn:vlp-energy-gain-condition}
      \left(l_\text{avg} - L(\theta)\right)sin(2\theta) \geq 0 
      \text{ }\forall \theta \in \mathbb{S}^1
      ,
   \end{equation}
   with the property that the inequality is strict except at the
   coordinate axes.
   If the inequality is flipped, the VNHC dissipates energy. 
\end{thm}
\begin{proof}
   The constraint manifold for \(h\) is the cylinder \(\Gamma = \mathbb{S}^1 \times \R \). 
   In negative-time, the dynamics on \(\Gamma\) are
   \begin{equation}\label{eqn:negative-time-vlp}
      f(q,p) = 
      -\begin{bmatrix}
         \dot{q} \\ \dot{p}
      \end{bmatrix} = \begin{bmatrix}
         -\frac{p}{mL(\theta)^2} \\
         mgL(\theta)\sin(q)
      \end{bmatrix}
      .
   \end{equation}
   Because \(q \in \mathbb{S}^1\), the only equilibria of these dynamics are 
   at \((q,p) = (0,0)\) and \((q,p) = (\pi,0)\).
   To prove the origin of the positive-time system (\ref{eqn:vlp-hamiltonian})
   is a repeller, we will show the origin of the negative-time system 
   (\ref{eqn:negative-time-vlp}) is almost everywhere globally asymptotically
   stable.

   Choose, as a candidate Lyapunov function, the energy for the
   average-length pendulum with zero-potential at the bottom of the swing:
   \[
       E_\text{avg}(q,p) := \frac{1}{2}\frac{p^2}{m l_\text{avg}^2} 
           + m g l_\text{avg} (1-\cos(q))
      .
   \]
   This is positive definite at \((0,0)\) and has compact sublevel
   sets on \(\Gamma\). 
   The derivative of \(E_\text{avg}\) along (\ref{eqn:vlp-hamiltonian}) is
   \begin{equation}\label{eqn:eavg-dot-pos}
      \dot{E}_\text{avg}^+ = -\frac{g\sin(q)p \left(L(\theta)^3 - l_\text{avg}^3\right)}
              {l_\text{avg}^2L(\theta)^2}
      ,
   \end{equation}
   and the derivative of \(E_\text{avg}\) along (\ref{eqn:negative-time-vlp}) is
   \begin{equation}\label{eqn:eavg-dot-neg}
      \dot{E}_\text{avg}^- = - \dot{E}_\text{avg}^+
      .
   \end{equation}
   The sign of \(\sin(q)p\) depends only on the quadrant in the
   \((q,p)\)-plane. In fact, it is easy to show that
   \[
      \sign{\sin(q)p} = \sign{\sin(2\theta)}
      .
   \]
   Furthermore, by Lemma \ref{lemma:sign-of-cube} we have
   \[ 
       \sign{L(\theta)^3 - l_\text{avg}^3} = \sign{L(\theta) - l_\text{avg}}
       .
   \]
   The positive-time derivative (\ref{eqn:eavg-dot-pos}) is nonnegative since
   \begin{align}\label{eqn:vlp-proof-neg-inv}
    \sign{\dot{E}_\text{avg}^+} &= \sign{-\sin(q)p \left(L(\theta)^3 - l_\text{avg}^3\right)}
          \nonumber \\
          &= \sign{-\sin(2\theta) \left(L(\theta) - l_\text{avg}\right)}
          \nonumber \\
          &= \sign{(l_\text{avg} - L(\theta))\sin(2\theta)} 
          \nonumber \\
          &\geq 0 \text{ (by assumption)}
      .
   \end{align}
   Using (\ref{eqn:eavg-dot-neg}) and (\ref{eqn:vlp-proof-neg-inv}) 
   we can determine that \(\sign{\dot{E}_\text{avg}^-} \leq 0\). 
   Thus, \(E_\text{avg}\) is a Lyapunov function for
   (\ref{eqn:negative-time-vlp}), which means the negative-time system is stable
   \cite{lyapunov}.

   Let us briefly analyze the upright equilibrium.
   Label the set of initial conditions converging to \((\pi, 0)\) in
   positive time by
   \[
      \Pi_0^+ := \left\{ (q_0,p_0) \in \Gamma \mid \lim_{t \to \infty} (q(t),p(t)) = (\pi, 0) \right\}
      ,
   \]
   and the set of initial conditions converging to \((\pi,0)\)
   in negative time by \(\Pi_0^-\).
   Take now the sublevel set
   \[
      E_{(\pi,0)} := \left\{ (q,p) \in \Gamma \mid E_\text{avg}(q,p) \leq
         E_\text{avg}(\pi,0) \right\}
      .
   \] 
   From (\ref{eqn:vlp-proof-neg-inv}), all sublevel sets of \(E_\text{avg}\) are
   negatively invariant in positive-time, which means
   complements of sublevel sets are positively invariant.
   Hence, the system cannot possibly converge to \((\pi,0)\) if it is
   initialized on 
   \(\left(E_{(\pi,0)}\right)^C = \Gamma \backslash E_{(\pi,0)}\).
   This implies that \(\Pi_0^+ \subset E_{(\pi,0)}\).
   Similarly, the interior of \(E_{(\pi,0)}\) is positively
   invariant for (\ref{eqn:negative-time-vlp}), 
   which implies \(\Pi_0^- \subset \left(E_{(\pi,0)}\right)^C \cup (\pi,0)\).

   The linearization of (\ref{eqn:vlp-hamiltonian}) at \((\pi, 0)\) is given by
   \begin{equation}\label{eqn:linearized-vlp}
      \begin{bmatrix}
         0 & \frac{1}{mL(0)^2} \\
         m g L(0) & 0 \\
      \end{bmatrix}
      ,
   \end{equation}
   which has eigenvalues at \(\lambda = \pm \sqrt{\frac{g}{L(0)}}\).
   Linearizing (\ref{eqn:negative-time-vlp}) at this point yields the same eigenvalues.
   Since one eigenvalue of (\ref{eqn:linearized-vlp})
   is positive and one is negative, the stable manifold theorem tells
   us that both \(\Pi_0^+\) and \(\Pi_0^-\) are one-dimensional
   \cite{stable_manifold}.
   This implies the origin cannot be globally asymptotically stable for
   (\ref{eqn:negative-time-vlp}); however, we can still show it is asymptotically stable
   for almost every initial condition because \(\Pi_0^+\) and \(\Pi_0^-\) are of
   measure zero within \(\Gamma\).

   We cannot linearize (\ref{eqn:vlp-hamiltonian}) at \((0,0)\) to prove 
   the origin is a repeller, because \(\theta(0,0)\) is not well-defined. 
   Instead, let \(D := \Gamma \backslash \Pi_0^-\).
   We will prove that (\ref{eqn:negative-time-vlp}) asymptotically converges to
   \((0,0)\) if it is initialized on \(D\).
   First, define the zero-set
   \[
      Z = \left\{(q,p) \in \Gamma \mid 
      \dot{E}_\text{avg}(q,p) = 0 \right\}
      .
   \]
   It is easy to show that \(Z\) is the union of the coordinate axes
   \[
      Z= \mathbb{S}^1 \times \left\{0\right\} \cup 
         \left\{0\right\} \times \R
      .
   \]
   Krasovskii-LaSalle tells us that solutions to (\ref{eqn:negative-time-vlp})
   converge to the largest invariant set in \(Z \cap D\)
   \cite{krasovskii_lasalle}.
   Since \(f(q,p) = (0,0)\) on \(D\) only when \((q,p) = (0,0)\),
   the origin is asymptotically stable on \(D\) for the negative-time system.
   In particular, the origin is asymptotically stable for every initial
   condition in the interior of \(E_{(\pi,0)}\).
   Hence, the positive-time system is diverging away from the origin 
   in the \((q,p)\) plane, which means the origin is a repeller.

   We now prove the system escapes compact sets in finite time.
   Let \(x(t) = (q(t),p(t))\) be the orbit for the positive-time system.
   Let \(K \subset \Gamma\) be compact and assume 
   \(x(0) \in K\backslash \left(\Pi_0^+ \cup \left\{(0,0)\right\}\right)\).
   For a proof by contradiction, suppose that for all \(T > 0\) there exists
   \(t' > T\) where \(x(t') \in K\).
   Let 
   \[
      k := \max\limits_{(q,p) \in K} E_\text{avg}(q,p)
      ,
   \] 
   be the largest energy value attained in \(K\), and let \(E_k\) be its
   sublevel set.

   Since \(E_k\) contains \(K\) and \(x(t') \in K\), it follows that 
   \(x(t') \in E_k\). 
   Furthermore, negative invariance of \(E_k\) means the half orbit
   \(x([0,t'])\) is contained fully within \(E_k\).
   In particular, \(x(T) \in E_k\).
   However, \(T\) was chosen arbitrarily, which implies \(x(t) \in E_k\) for all
   \(t \in [0,\infty[\).
   Since \(E_k\) is compact, \(x(t)\) has a positive limit set in
   \(E_k\) which, by Poincar\'{e}-Bendixson, is either an equilibrium or a
   closed orbit \cite{poincare_bendixson}.
   We will show this positive limit set must be an equilibrium and thereby
   reach our contradiction.

   Suppose the system had a closed orbit. 
   The orbit would necessarily be contained in a
   level set of \(E_\text{avg}\), since this is a non-decreasing function and
   closed orbits are periodic.
   Thus, \(E_\text{avg}\) is constant along the orbit, which means the
   orbit is an invariant subset of the set \(Z\) consisting of the two axes.
   However, the only invariant subsets of \(Z\) are equilibria, so 
   the system cannot contain closed orbits.

   The lack of closed orbits means the positive limit set of \(x(t)\) must 
   be one of the equilibria. 
   The origin is a repeller and \(x(0) \neq (0,0)\), which means \(x(t)\) is
   converging to \((\pi,0)\).
   This is a contradiction, since we assumed \(x(0) \not \in \Pi_0^+\).

   We conclude that the VNHC is injecting energy into the VLP.
   By flipping the inequality of (\ref{eqn:vlp-energy-gain-condition})
   we find the VLP is gaining energy in negative-time, so the VNHC is
   dissipating energy.
\end{proof}

>>>>

Using this configuration, we will compute the Hamiltonian dynamics of the system.
The Cartesian position of the mass at the tip of the pendulum
is given by \(x = (l\sin(q),-l\cos(q))\), while its velocity is
\(\dot{x} = (\dot{l}\sin(q) + l\cos(q)\dot{q}, -\dot{l}\cos(q) + l\sin(q)\dot{q})\).
Computing the kinetic energy \(T\) yields
\[
   T(\mathbf{q},\dot{\mathbf{q}}) = 
   \frac{1}{2}m\norm{\dot{x}}^2 = \frac{1}{2}m\left(\dot{l}^2 + l^2\dot{q}^2\right)
   .
\]
The potential energy \(P\) with respect to the pivot (under a gravitational
acceleration \(g\)) is
\[
   P(\mathbf{q}) = -mgl\cos(q)
   .
\]
Collecting the kinetic energy into a quadratic form, we get the Lagrangian
\[
   \mathcal{L}(\mathbf{q},\dot{\mathbf{q}}) 
   = \frac{1}{2} \dot{\mathbf{q}}\tpose D(\mathbf{q})\dot{\mathbf{q}} - P(\mathbf{q})
   = \frac{1}{2}
   \begin{bmatrix} \dot{q} & \dot{l} \end{bmatrix}
   \begin{bmatrix}
      ml^2 & 0 \\
      0 & m \\
   \end{bmatrix}
   \begin{bmatrix} 
      \dot{q} \\ \dot{l}
   \end{bmatrix}
   + mgl\cos(q)
   .
\]
Computing the conjugate of momenta to \(\mathbf{q}\), we get 
\[
   \mathbf{p} := \begin{bmatrix} p \\ p_l \end{bmatrix} 
   = \begin{bmatrix} ml^2\dot{q} \\ m\dot{l} \end{bmatrix} 
   .
\]
Performing the Legendre transform on \(\mathcal{L}\) and setting
\(M(\mathbf{q}) := D(\mathbf{q})\), \(V(\mathbf{q}) := P(\mathbf{q})\),
we get the Hamiltonian (\ref{eqn:simply-actuated-hamiltonian}) whose
dynamics (\ref{eqn:simply-actuated-dynamics}) resolve to  
\begin{align}\label{eqn:vlp-hamiltonian-with-pl}
   \mathcal{H} &= \frac{1}{2} \begin{bmatrix} p & p_l \end{bmatrix}
      \begin{bmatrix}
         \frac{1}{ml^2}  & 0 \\
         0 & \frac{1}{m}
      \end{bmatrix} \begin{bmatrix} p \\ p_l \end{bmatrix} - mgl\cos(q)
      , \\
     &\begin{cases}
        \dot{q} = \frac{p}{ml^2} \\
        \dot{l} = \frac{p_l}{m} \\
        \dot{p} = -mgl\sin(q) \\
        \dot{p}_l = \frac{p^2}{ml^3} + mg\cos(q) + \tau
        . \\
   \end{cases} \nonumber
\end{align}
The control input is a force \(\tau \in \mathbb{R}\) affecting the dynamics of
\(p_l\), acting collinearly with the rod.
We assume the force does not affect the dynamics of \(p\) in any way -
that is, the control input cannot enact any lateral force on the pendulum.
This makes the VLP into an underactuated mechanical system with degree of
underactuation one. 
It is also a useful assumption because it means \((\mathbf{q},\mathbf{p})\) 
are simply actuated coordinates, which allows us to apply the theory of VNHCs we
developed in Chapter \ref{ch:vnhcs}.

@@

\begin{defn}\label{defn:energy-injection}
   A regular VNHC \(h(\mathbf{q},\mathbf{p}) = 0\) \textit{injects (dissipates)
     energy} if the constrained dynamics in~\eqref{eqn:vlp-hamiltonian} gain
   (lose) energy everywhere on the constraint manifold, except possibly on a set
   of measure zero.
\end{defn}

@@

\begin{thm}\label{thm:vlp-energy-stabilization}
   Define \(\theta := \arctan_2(p,q)\).
   Let \(L : \mathbb{S}^1 \rightarrow [\underline{l},\overline{l}]\) be a
   \(C^2\) function of \(\theta\).
   A regular VNHC of the form 
   \(h(\mathbf{q},\mathbf{p}) = l - L(\theta)\) injects energy into the VLP
   if there exists 
   \(l_\text{avg} \in [\underline{l},\overline{l}]\) such that 
   \begin{equation}\label{eqn:vlp-energy-gain-condition}
      \left(l_\text{avg} - L(\theta)\right)sin(2\theta) \geq 0 
      \text{ }\forall \theta \in \mathbb{S}^1
      ,
   \end{equation}
   with the property that the inequality is strict except at $\theta \in
   \{0,\pi/2,\pi,3 \pi/2\}$.  If instead 
\[
 \left(l_\text{avg} - L(\theta)\right)sin(2\theta) \leq 0 
      \text{ }\forall \theta \in \mathbb{S}^1
      ,
\]
the VNHC dissipates energy.
\end{thm}
\begin{proof}
   We will first show that, under the given assumptions, the origin of the
   constrained dynamics~\eqref{eqn:vlp-hamiltonian} is a repeller, and for that
   we consider the negative time system obtained
   from~\eqref{eqn:vlp-hamiltonian} by reversing the sign of the vector field:
\begin{equation}\label{eqn:negative-time-vlp}
   \begin{cases}
      \dot{q} = -\frac{p}{m L^2} \\
      \dot{p} = mgL\sin(q)
      . \\ 
   \end{cases}
\end{equation}
The state space of~\eqref{eqn:negative-time-vlp} is \(\mathbb{S}^1
\times \R\), and the system has two equilibria, \((q,p)=(0,0)\) and
\((q,p)=(\pi,0)\). We need to show that the equilibrium
\((q,p)=(0,0)\) is asymptotically stable
for~\eqref{eqn:negative-time-vlp}.

Consider the Lyapunov function candidate
   \[
       E_\text{avg}(q,p) := \frac{1}{2}\frac{p^2}{m l_\text{avg}^2} 
           + m g l_\text{avg} (1-\cos(q))
      ,
   \]
which corresponds to the energy of VLP when the pendulum length is fixed at
\(l_\text{avg}\).  The function \(E_\text{avg}\) is positive definite at
\((0,0)\), and has compact sublevel sets on \(\mathbb{S}^1 \times \R\).  The
derivative of \(E_\text{avg}\) along~\eqref{eqn:negative-time-vlp} is
   %% \begin{equation}\label{eqn:eavg-dot-pos}
   %%    \dot{E}_\text{avg}^+ = -\frac{g\sin(q)p \left(L(\theta)^3 - l_\text{avg}^3\right)}
   %%            {l_\text{avg}^2L(\theta)^2}
   %%    ,
   %% \end{equation}
   %% and the derivative of \(E_\text{avg}\) along (\ref{eqn:negative-time-vlp}) is
   \begin{equation}\label{eqn:eavg-dot-neg}
      \dot{E}_\text{avg} = \frac{g\sin(q)p \left(L(\theta)^3 -
        l_\text{avg}^3\right)} {l_\text{avg}^2L(\theta)^2}
      .
   \end{equation}
It is easy to show that
   \[
      \sign{\sin(q)p} = \sign{\sin(2\theta)}
      .
   \]
   Furthermore, by Lemma \ref{lemma:sign-of-cube} we have
   \[ 
       \sign{L(\theta)^3 - l_\text{avg}^3} = \sign{L(\theta) - l_\text{avg}}
       ,
   \]
and therefore,
   \begin{align}\label{eqn:vlp-proof-neg-inv}
    \sign{\dot{E}_\text{avg}} &= \sign{\sin(q)p \left(L(\theta)^3 - l_\text{avg}^3\right)}
          \nonumber \\
          &= \sign{\sin(2\theta) \left(L(\theta) - l_\text{avg}\right)}
          \nonumber \\
          &= -\sign{(l_\text{avg} - L(\theta))\sin(2\theta)} 
          \nonumber \\
          &\leq 0 \text{ (by assumption)}
      .
   \end{align}
We have thus shown that \(\dot{E}_\text{avg} \leq 0\), and therefore
the equilibrium \((q,p)=(0,0)\) is stable for~\eqref{eqn:negative-time-vlp}. We
now apply the Krasovskii-LaSalle invariance principle, and consider the largest
subset of \( Z = \left\{(q,p) \in \mathbb{S}^1 \times \R \mid
\dot{E}_\text{avg}(q,p) = 0 \right\} \). We see from~\eqref{eqn:eavg-dot-neg}
that \(\dot{E}_\text{avg}(q,p)\) is zero when $p=0$, \(\sin(q)=0\), or
$L(\theta) = l_\text{avg}$. This latter condition by assumption is met when
\(\theta \in \{0,\pi/2,\pi, 3\pi/2\}\), or equivalently when either $q=0$ or
$p=0$. All in all, we have that
   \[
      Z= \left\{(q,p) \in \mathbb{S}^1 \times \R \mid q =0\right\} \cup \left\{(q,p) \in \mathbb{S}^1 \times \R \mid p =0\right\} 
      .
   \]
It is easily seen that the largest invariant subset of \(Z\) is the union of the two equilibria
\[
\Omega =  \{(0,0)\}\cup \{(\pi,0)\}
.
\]  
Since all sublevel sets of \(E_\text{avg}\) are compact, the
Krasovskii-LaSalle invariance principle implies that the set
\(\Omega\) is globally attractive. Since the two equilibria are
isolated, and since \((q,p)=(0,0)\) is stable, we deduce that the
equilibrium \((q,p)=(0,0)\) is asymptotically stable for the negative
time system~\eqref{eqn:negative-time-vlp}, and thus it is a repeller
for the constrained dynamics in~\eqref{eqn:vlp-hamiltonian}.

We now prove that almost all solutions of the reduced
dynamics~\eqref{eqn:vlp-hamiltonian} escape compact sets in finite
time, i.e., that for each compact subset \(K\) of \(\mathbb{S}^1
\times \R\), and for almost every condition \( (q_0,p_0) \in K \),
there exists \( T > 0\) such that \( (q(t),p(t)) \not \in K \) for all
\( t > T \), where \( (q(t),p(t)) \) denotes the solution
of~\eqref{eqn:vlp-hamiltonian} with \( (q(0),p(0)) = (q_0,p_0)\). In
what follows, we denote by \( x = (q,p) \) the state of the reduced
dynamics~\eqref{eqn:vlp-hamiltonian}. Also, we denote by \(\Pi^-\) the
stable manifold of the equilibrium \((\pi,0)\) for
system~\eqref{eqn:vlp-hamiltonian}, defined as
 \[
      \Pi^- := \left\{ x(0) \in \mathbb{S}^1 \times \R \mid
      \lim_{t \to \infty} x(t) = (\pi, 0) \right\}.
   \]
The linearization of system~\eqref{eqn:vlp-hamiltonian} at the
equilibrium \((\pi,0)\) has a system matrix given by
  \begin{equation}\label{eqn:linearized-vlp}
      \begin{bmatrix}
         0 & -\frac{1}{mL(0)^2} \\
         -m g L(0) & 0 \\
      \end{bmatrix}
      ,
   \end{equation}
whose spectrum is \(\{\pm \sqrt{{g}/{L(0)}}\}\).  Since one eigenvalue
of (\ref{eqn:linearized-vlp}) is positive and one is negative, the
stable manifold theorem implies that the stable manifold \( \Pi^- \)
is a one-dimensional immersed submanifold of \(\mathbb{S}^1 \times
\R\), and thus a set of zero measure.

Let \( K \subset \mathbb{S}^1 \times \R\) be an arbitrary compact set
and let \( x(0) \in K \backslash \left(\Pi^- \cup
\left\{(0,0)\right\}\right)\). Since \( x(0) \not \in \Pi^- \), we have that
\begin{equation}\label{eqn:vlp-x-does-not-converge}
x(t) \not \to_{t \to \infty} (\pi,0).
\end{equation}
Suppose, by way of contradiction, that for each \(T
> 0\) there exists \(t' > T\) such that \(x(t') \in K\).  Letting
   \[
      k := \max\limits_{x \in K} E_\text{avg}(x)
      ,
   \] 
and \(E_k =\{ x \mid E_\text{avg}(x) \leq k\} \), we have that \( K
\subset E_k \), and thus \(x(t') \in E_k\). We have shown earlier that
the function \( E_\text{avg} \) is nonincreasing along solutions
of~\eqref{eqn:negative-time-vlp}, which implies that \(
E_\text{avg}(x(t)) \) is nondecreasing for solutions of the reduced
dynamics~\eqref{eqn:vlp-hamiltonian}, and therefore the half orbit
\(x([0,t'])\) is contained fully within \(E_k\).  In particular,
\(x(T) \in E_k\). Since this is true for each \( T >0 \), we deduce
that \(x(t) \in E_k\) for all \(t \in [0,\infty[\).  Since \(E_k\) is
    compact, \(x(t)\) has a positive limit set in \(E_k\) which, by
    the Poincar\'{e}-Bendixson theorem, is either an equilibrium or a
    closed orbit \cite{poincare_bendixson}.  We will show this
    positive limit set must be an equilibrium and thereby reach a
    contradiction.

In order to rule out closed orbits, suppose that the reduced
dynamics~\eqref{eqn:vlp-hamiltonian} have a periodic solution \(
z(t)\). Since \( E_\text{avg}(z(t)) \) is nondecreasing and \(z(t)\)
is periodic, it must be that \( E_\text{avg}(z(t)) \) is constant, or
\(\dot E_\text{avg}(z(t)) \equiv 0\). We have shown earlier
that\footnote{We have shown it for the negative time
  system~\eqref{eqn:negative-time-vlp}, but changing the sign of the
  vector field does not change the set where \(\dot E_\text{avg}
  =0\) nor the invariant sets.}
\[
\{x \mid \dot E_\text{avg}(x)=0\} = \left\{(q,p) \mid q =0\right\}
\cup \left\{(q,p) \mid p =0\right\},
\]
and the only invariant subset of this set is the union of two disjoint
equilibria. Since the orbit \(z(\R)\) is an invariant set which is
connected, \(z(\R)\) must be an equilibrium and cannot be a nontrivial
closed orbit.

Returning to the Poincar\`e-Bendixson theorem, the positive limit set
of \( x(t)\) must be an equilibrium and,
by~\eqref{eqn:vlp-x-does-not-converge}, this equilibrium must be \(
(0,0)\). Since we have chosen \(x(0) \neq (0,0)\), and since \( (q,p) = (0,0) \) is a
repeller, \( x(t) \) cannot converge to \( (0,0)\), which gives a
contradiction.

We conclude that the VNHC injects energy into the VLP.
   By flipping the inequality of (\ref{eqn:vlp-energy-gain-condition})
   we find the VLP is gaining energy in negative-time, so the VNHC is
   dissipating energy.
\end{proof}
